DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-06-28 manuscript/manuscript.tex
add updated files for review submission

School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.

@@

Word count: 6715

@@

Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data, also in a computer-readable format (Stoltzfus et al., 2013); and methods to summarize and compare the data.

@@

At the time of writing of this manuscript
(May 15, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009).
If the source chronograms do not form a grove, the supertree reconstruction will fail.
In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.

@@

\textbf{\emph{Dating a tree with no branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.

@@

We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study. On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3. As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy. The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.

@@

\textbf{\emph{Creating a search query.--}}
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3A, which would not have had any data otherwise.

\textbf{\emph{Searching the database.--}}
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig. 3B). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table 1).

\textbf{\emph{Summarizing search results.--}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. 3A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. 3C4).

@@

\textbf{\emph{Searching the database.--}}
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).

\textbf{\emph{Summarizing search results.--}}
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. 5). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).

Age data from source chronograms was congruified to OpenTree's topology (Fig. 5B), reducing the age data set to 818 different data points (Supplementary Table S2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. 4).

@@

We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

@@

As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results. This \enquote{garbage in, garbage out} problem has long been recognised in supertree methods for summarizing phylogenetic trees (Bininda-Emonds et al., 2004).
We note that this is a surfacing issue of any automated tool for biological data analysis. For example, DNA riddled with sequencing errors will produce generally poor alignments that will return biased evolutionary hypothesis, independently of the quality of the analysis software used. Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.
Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others. For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality (Webb et al., 2008). In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs. 4, 5; Supplementary Figs. S1, S5).
Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence (Sauquet, 2013).
Finally, placement of calibrations also affects uncertainty of node age estimates. For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals (Vos \& Mooers, 2004), which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

@@

While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis (Antonelli et al., 2017; Garzón-Orduña, Silva-Brandão, Willmott, Freitas, \& Brower, 2015; Graur \& Martin, 2004; Sauquet, 2013; Sauquet et al., 2012; Schenk, 2016; Shaul \& Graur, 2002).
However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply calibrations appropriately (Hipsley \& Müller, 2014; Ksepka et al., 2011). One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees (Ksepka et al., 2011; Sauquet, 2013), and making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al. (2017).
The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications (Ksepka et al., 2015). This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.

In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record). Recently, Powell, Waskin, and Battistuzzi (2020) showed in a simulation analysis that secondary calibrations are as good as primary calibrations. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before (Graur \& Martin, 2004; Sauquet, 2013; Shaul \& Graur, 2002). Our cross validation analysis also provides some insight into the application of secondary calibrations. In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs. S2-S20). We also note that nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies. The only exception to this trend was observed in Burns et al. (2014) chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations (Sauquet et al., 2012; Schenk, 2016).

We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002). It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003); with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005); and with methods that summarize age data unto a tree topology, such as DateLife.

@@

Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data. These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).

@@

The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
It is available as an R package, and as a web-based R shiny application at www.datelife.org.
Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research. The DateLife project will improve awareness of the existing variation in expert estimations time of divergence data, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

@@

Supplementary Figures can be viewed and dowloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}). Development versions corresponding to all of the above are hosted on GitHub, accesible at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.

@@

\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.

>>>>

Department of Life and Environmental Sciences, University of California, Merced, CA 95343, USA.

@@

Word count: 6707

@@

Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to a wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data, also in a computer-readable format (Stoltzfus et al., 2013); and methods to summarize and compare the data.

@@

At the time of writing of this manuscript
(Jun 11, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1,239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009).
If the source chronograms do not form a grove, the supertree reconstruction will fail.
In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.

@@

\textbf{\emph{Dating a tree without branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will choose an age for the root so that it can return a dated topology. It will also provide a conspicuous warning message so that users are aware that the root of the chronogram does not have information available in the chronogram database, along with suggestions on how the user can provide an appropriate age for the root.

@@

We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study. On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3. As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy. The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.

@@

\textbf{\emph{Creating a search query.--}}
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3a, which would not have had any data otherwise.

\textbf{\emph{Searching the database.--}}
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig. 3b). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table 1).

\textbf{\emph{Summarizing search results.--}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. 3A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
Median summary age data for node \enquote{n2} was excluded as final calibration because it is older than the median age of a more inclusive node, \enquote{n1} (Fig. 3c4).

@@

\textbf{\emph{Searching the database.--}}
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1,206 different age data points available for species within the Fringillidae (Supplementary Table S1).

\textbf{\emph{Summarizing search results.--}}
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. 4a). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Fig. 4b; Open Tree Of Life et al., 2019).

Source chronograms (Supplementary Figs. S2-S20) were congruified to OpenTree's topology shown in Figure 4b, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that can be used as calibrations for the chosen topology (Fig. 4b). The congruent node age data points were summarized for each node, resulting in 194 summary node ages. From these 21 were excluded as secondary calibrations because they were older than the ancestral node. The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig. 5).

@@

We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 6).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

@@

As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results. The \enquote{garbage in, garbage out} problem has long been recognised in supertree methods for summarizing phylogenetic trees (Bininda-Emonds et al., 2004).
We note that this is a surfacing issue of any automated tool for biological data analysis. For example, DNA riddled with sequencing errors will produce generally poor alignments that will return biased evolutionary hypothesis, independently of the quality of the analysis software used. Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.

Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others. For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality (Webb et al., 2008). In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs. 5, 6; Supplementary Figs. S1c, S4).
Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence (Sauquet, 2013).
Finally, placement of calibrations also affects uncertainty of node age estimates. For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals (Vos \& Mooers, 2004), which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

@@

DateLife constructs summary chronograms using node ages extracted from existing chronograms, i.e.~secondary calibrations.
In general, the scientific community has more confidence in chronograms using primary calibrations, where the dated tree is generated from a single analysis where carefully chosen fossil calibrations are the source of absolute time information, than in analyses dated using secondary calibrations (Antonelli et al., 2017; Garzón-Orduña, Silva-Brandão, Willmott, Freitas, \& Brower, 2015; Graur \& Martin, 2004; Sauquet, 2013; Sauquet et al., 2012; Schenk, 2016; Shaul \& Graur, 2002).
However, implementation of primary calibrations is difficult: it requires specialized expertise and training to discover, place and apply calibrations appropriately (Hipsley \& Müller, 2014; Ksepka et al., 2011). One approach is to use fossils that have been widely discussed and previously curated as calibrations to date other trees (Ksepka et al., 2011; Sauquet, 2013), and making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al. (2017).
The Fossil Calibration Database provides data for 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewed publications (Ksepka et al., 2015). This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
Yet, users still require the expertise to locate and calibrate appropriate nodes in their phylogenies which correspond with fossils available in the database.

Recently, Powell, Waskin, and Battistuzzi (2020) showed in a simulation study that secondary calibrations using node ages based on previous molecular clock analyses can be as good as primary calibrations.
Using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases (Graur \& Martin, 2004; Sauquet, 2013; Shaul \& Graur, 2002). Our cross validation analysis also provides insight into the application of secondary calibrations. Node ages summarized with DateLife and those from the original studies are well correlated (Supplementary Figs. S2-S20). We also note that DateLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using our secondary calibrations than ages from the original studies. The only exception to this trend was observed in Burns et al. (2014) chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations (Sauquet et al., 2012; Schenk, 2016). However, these younger dates are more likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the Burns et al. (2014) tree were obtained using a single secondary calibration based on a previously estimated molecular evolution rate across birds from Weir and Schluter (2008), and appear as major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (Fig. 5).

Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002).

@@

Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353,185 seed plants where only 23\% had molecular data. These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).

@@

The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
It is available as an R package, and as a web-based R shiny application at www.datelife.org.
DateLife provides a straightforward way to get an informed picture of the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
Additionally, both summary and newly generated trees using the DateLife workflow are useful to evaluate evolutionary hypotheses in different areas of research. We hope that the DateLife project will increase awareness of the existing variation in expert estimations of time of divergence, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

@@

Supplementary Figures can be viewed and downloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}). Development versions corresponding to all of the above are hosted on GitHub, accesible at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.

@@
